Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-06-03 analysis/paper/paper2.Rmd
Re-writes

# Introduction
Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access data and reproducible research. To begin with the first of these, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention. The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.

For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA). Research into this once maligned period has grown apace in the last few decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [on the origins of which, see @Kotsonas2016; @Morris1997]. Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.

Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a number of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period. This is reflected in survey work, which has focussed extensively around the Bay of Mirabello, where such sites are numerous [@Haggis2005; @Hayden2005; @Watrous2012]. This has, unfortunately, engendered a certain disregard for settlement characterised by different topographies, especially in the centre of the island. There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across Crete in the EIA [@Kotsonas2011b; @Whitley2011, 667–8].

As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean. Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative research [@Bevan2013; @Knappett2011a; @Knappett2011]. The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011]. The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and have been studied with regard to their 'oriental' imports [@Antoniadis2012], antique artefacts [@Crowe2016], and religious context [@Coldstream1984b]. But a lack of quantitative analyses, or systematic comparisons between the cemeteries, has left us with many intuited trends, some exceptional tombs, but a general impression of un-patterned multiplicity.

With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis seeks to address this present lack. In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer not just generalities, but tangible and context-specific considerations of the motivations and identities behind the burial groups themselves, with a view to eventually comparing such results with mortuary assemblages across the island.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991, 287]. Most individual finds of Protogeometric or Geometric date take the form of flimsy foundations, patches of earth floor, wells or simple scatters of pottery, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].

However, a concentration of activity slightly to the west of the old palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], while, more recently, findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha [@Kotsonas2011a, 5-8]. This estimate remains our best guess in the absence of substantial excavated deposits but, if it is accurate, then EIA Knossos was one of the largest sites of the contemporary Aegean.

@@

Survey aside, our most abundant evidence for EIA Knossos comes from the mortuary record (see Figure \@ref(fig:map). The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with multiple burials known from the Kephala ridge [@Coldstream1963, 38; @Hogarth1899, 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968, 133–46], the cave of Mavro Spelio [@Antoniadis2012, 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997 @Kotsonas2006]. Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed and often isolated that they provide little scope for systematic quantitative analysis.

Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods. Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary practices, but for social life of any kind at EIA Knossos.

@@

The cemetery on the slope facing the eponymous village of Fortetsa is composed of twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi). These tombs cluster in three main groups. The largest of these, comprising nine tombs, is arranged on a north-south axis, with the *dromos* of each tomb extending downslope to the west. To the north lies a cluster of five tombs and, to the south, another of three. Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known, after the year of their excavation, as the Fortetsa 1967 tombs, were probably the southernmost burials of the KNC. Unfortunately, all had been looted at the time of excavation, and no *in situ* deposit was found in any tomb. They are thus excluded from present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC. The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations here in 1978 [@Coldstream1996a: 53-55].

The tombs themselves fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-8]. The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted to child burials and [@Antoniadis2012: 69].

@@

The initial foundation of the cemeteries has received much attention. It has been claimed that no tomb crosses the Late Minoan (LM) IIIC-SM divide [@Coldstream1998, 58; @Coldstream2006, 582], though there is a possibility that this is a classificatory artefact [@Hallager2010]. Instances of BA tombs reused in the EIA are known (e.g. the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though other evidence would suggest this is unlikely (Whitelaw, pers. comm.).

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment. The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957, 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b, 93-94]. Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T. 107; see @Coldstream1984b; -@Coldstream1994, 112–13]. Whether simply nostalgia [@Coldstream1998, 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994, 114-15], it seems that the Bronze Age past possessed a significant cultural cachet mobilised by various EIA burying groups.

The final mystery of the chronological sequence at the KNC and Fortetsa is the quite abrupt cessation of burials around 630 BC. There had been changes in tomb use in both cemeteries in the preceding century, with few new tombs built, though interments in old tombs continued [@Brock1957, 4; @Cavanagh1996; 651-3]. This curiosity has, along with the equally scanty evidence for the subsequent period in the wider settlement, become known as the ‘Archaic Gap'. The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown. Although various explanations have been forwarded, [@Coldstream1999, 301-2; @Huxley1994, 126; @Kotsonas2002, 41-4], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers comm.).

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’, a term derived from Appadurai [-@Appadurai1986] and Kopytoff [-@Kopytoff1986]. Influenced by Mauss’ [-@Mauss1954] formulation of ‘the gift’, they stressed the mutuable nature of an object’s value, and its capacity for acquiring a ‘biography’ through the transformations of its social existence. These ideas, fertilised by the Homeric motif of gift-giving [@Whitley2002, 220-1], and the remarkable archaeological parallels for such objects – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - have been brought to bear on certain elaborate burials of the EIA.

Catling’s discussion of the Tomb 200+ complex is a notable example. Tomb 201 contained the remains of at least two adults, and possibly a child, a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995, 123]. Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi. All contained imported items, were of ‘warrior grave’ type, and formed loci for later burials. Catling argues that those individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment in death [@Catling1995, 127–8].

Complementing the theme of heroes abroad has been a similar interest in foreigners at home. Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman. Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006, 155–9]. While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997, 176–85; @Schreiber2003, 293-306].

@@

Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches. Cluster analyses have been employed by both Cavanagh [-@Coldstream1996, 653-7] and Antoniadis [-@Antoniadis2012, 193-7], the most sophisticated statistical techniques so far applied to the material. Antoniadis [-@Antoniadis2012, 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods. Yet even here the available data could be further interrogated, such as with the number of imported items *per burial*, something Antoniadis does not explore. The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Each entry includes, among other things, the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were accordingly categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe (2016). In total, this database comprises just over 4540 individual artefacts.

@@

Taken together, these databases constitute a detailed record of all the finds from the Fortetsa and KNC publications, and offer the opportunity for carrying out a range of quantitative analyses. It must be conceded that, with many of the tombs evidently looted or damaged over the centuries, this cannot be a complete picture of the tombs’ original contents. If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless. As Snodgrass [-@Snodgrass1996] notes, the situation means we should be extremely cautious about making negative arguments – that is, conclusions drawn about the lack of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.

@@

Evermore archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data underpinning our academic output, but the full methodologies by which our analyses are conducted, in formats facilitating reproducibilty. It is felt that, with the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article]. In Marwick's words [-@Marwick2017, 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".

@@

There are, of course, complications to the sharing of data and methods. There are valid reservations around issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research. But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit greatly from the greater accessability of datasets and workflows.

In this vein, the present article has been composed making use of Ben Marwick's "rrtools" [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio, which provides the author with the tools to create a self-contained "compendium", which can be freely downloaded, and from which anyone can accurately reproduce the entire workflow of the present analysis. Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R. This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and article as originally written are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**). The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or perceived errors or inaccuracies in the data. It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.

@@

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were sometimes used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)). In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.

@@

A central concern of the present study was how to deal with the temporal uncertainty implicit in the assigning of individual vessels to stylistically-defined periods. A vessel dated to the PGB period could, on our best estimates, date anywhere between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data through means which normally require values and not ranges. A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper when the division of vessels into centuries renders the inaccuracies of such a method less significant.

@@

But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi), a simple plot of vessel counts based on these average dates; the apparent cessation of depositions for 30 years in the 8th century is simply a fiction generated by the processing of the data.

A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis" [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than assign a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades. These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures. The major contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].

@@

As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations (see @Crema2010, 1123-1124), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.

@@

>>>>

# Introduction
Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access, reproducible research. To begin with the first, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention. The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.

For one, though, this is not the Knossos of that illustrous, palatial period, but rather the town as it persisted and evolved during the subsequent Early Iron Age (EIA). Research into this once maligned period has grown apace in recent decades, beginning with the pioneering works of Coldstream [-@Coldstream1977], Desborough [-@Desborough1972], and Snodgrass [-@Snodgrass1971a] and culminating in the rejection by most scholars of the appelation and interpretive bagguage of the so-called ‘Dark Ages’ [see @Kotsonas2016; @Morris1997]. Nonetheless, our knowledge of many EIA settlements remains often underdeveloped in comparison with those of the second millennium BC.

Secondly, and welcome though they are, a number of recent publications on the socio-economic [@Wallace2010a] and settlement [@Nowicki2000] changes which accompanied the BA-EIA transition have directed scholarly attention toward a pattern of elevated, defensible settlements supposedly founded in response to the dangers of this turbulent period. This is reflected in survey work, which has focussed extensively around the Bay of Mirabello, where such sites are numerous [@Haggis2005; @Hayden2005; @Watrous2012]. This has, unfortunately, engendered a certain disregard for settlement characterised by different topographies, especially in the centre of the island. There remains, therefore, a need for studies addressing the social dynamics at such sites (Knossos, Phaistos, Grivila, Eleutherna etc.), with a long-term view to comparative appraisal of the divergent trajectories of communities across Crete in the EIA [@Kotsonas2011b; @Whitley2011, 667–68].

As to the second and third of my stated interests, several authors have recently demonstrated the great potential of quantitative, statistical, spatial, and network analyses to elucidate social dynamics, settlement patterning, and systems of inter-regional communication across Crete and the wider Aegean. Such work encourages the use of large (often pre-existing) datasets, and the sharing of both data and methods to facilitate reproducible, collaborative research [@Bevan2013; @Knappett2011a; @Knappett2011]. The cemeteries of Knossos offer an opportunity to introduce such approaches into EIA research on Crete, where they have yet to make much of an impact [see @Kotsonas2011]. The Fortesta and North Cemetery complexes are both well-published [@Brock1957; @Coldstream1996a] and studied, but a lack of systematic, quantitative comparisons between the cemeteries has left us with many intuited trends, some exceptional tombs, yet a general impression of un-patterned multiplicity.

With a dataset including information on every find and tomb recorded in the Fortetsa and Knossos North Cemetery publications [@Brock1957; @Coldstream1996a], the following analysis addresses this present lack. In doing so, I aim to demonstrate the potential for quantitative methods, firstly, to permit identification, and nuanced characterisation, of significant patterning in the archaeological material; secondly, to facilitate a more systematic appraisal of the cemeteries’ largest tombs and their distinctive histories; and, finally, to offer tangible and context-specific evidence as to the behaviours and identities of the burying groups, with a view to eventually comparing such results with mortuary assemblages across Crete.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that “the remains of [this] period are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced into a single city, as per Aristotle’s model of synoecism [@1950].

However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8]. This estimate remains our best guess in the absence of substantial excavated deposits but, if accurate, suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.

@@

Survey aside, our most abundant evidence comes from the mortuary record (see Figure \@ref(fig:map)). The transitional period between the Late Bronze Age (LBA) and EIA yields few if any securely datable interments, but later evidence is more plentiful, with burials known from the Kephala ridge [@Coldstream1963: 38; @Hogarth1899: 82-5], the modern suburbs of Ayios Ioannis and Atsalenio [@Boardman1960a; @Davaras1968: 133–46], the cave of Mavro Spelio [@Antoniadis2012: 58-9], and the Khaniale Teke site with its well-known tholos [@Boardman1967;@Hutchinson1954; cf. @Hoffman1997 @Kotsonas2006]. Taken together, these scattered tombs point to a richly variegated set of funerary practices but, despite their reasonable number, are so dispersed that they provide little scope for systematic quantitative analysis.

Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods. Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.

@@

The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi). These cluster in three main groups. The largest of these contains nine tombs, arranged on a north-south axis, with the *dromos* of each extending downslope to the west. To the north lies a cluster of five tombs and, to the south, another of three. Finally, three tombs excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).

To the northeast of the Fortetsa cemetery, a collection of ten badly damaged tombs known as the Fortetsa 1967 tombs were probably the southernmost burials of the KNC. Unfortunately, all had been thoroughly looted at the time of excavation, and are thus excluded from present analysis. Northwest of these, and just south of the Teke Tholos, the Teke tombs represent the northern reaches of the KNC. The central section of the KNC comprises the ‘Medical Faculty’ site, named for the extension of the University of Crete that prompted major rescue excavations there in 1978 [@Coldstream1996a: 53-55].

The tombs themselves fall into three main types, the chamber tomb, shaft grave, and pit cave or tomb; the latter two are largely Subminoam phenomena, with the former by far the most abundant. It comprises a descending ramp or passage, the *dromos*, leading to the chamber itself, cut into the *kouskouras* bedrock. The *dromos* and chamber are separated by the *stomion*, an opening often set a step or two lower than the *dromos*, which is covered by a large stone slab, smaller stacked stones, or combination of both [@Antoniadis2012: 47-48]. The majority of burials are cremations, a rite that appears at Knossos in the Subminoan period, in the Tomb 200+ complex in the KNC [@Coldstream1994: 109]. Inhumation does not completely disappear in the EIA, but by the Orientalising period it appears restricted solely to child burials [@Antoniadis2012: 69].

@@

The foundation of the cemeteries has received much attention. It has been claimed that no tomb crosses the Late Minoan IIIC-SM divide [@Coldstream1998: 58; @Coldstream2006: 582], though there is a possibility that this is a classificatory artefact [@Hallager2010]. Instances of BA tombs reused in the EIA are known (e.g. the Khaniale Teke tholos), and Cavanagh [-@Cavanagh1996], based on a cluster analysis of tomb dimensions, could not reject the possibility of some KNC tombs being of BA date, though other evidence would suggest this is unlikely (Whitelaw, pers. comm.).

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment. The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957: 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b: 93-94]. Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T. 107; see @Coldstream1984b; -@Coldstream1994: 112–13]. Whether simply nostalgia [@Coldstream1998: 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994: 114-15], the Bronze Age past clearly possessed a significant and enduring cultural cachet for the EIA Knossians.

As contentious as their founding is the abrupt abandonment of both cemeteries around 630 BC. Admittedly, few new tombs had been built in the preceding century, though interments in old tombs had certainly continued [@Brock1957: 4; @Cavanagh1996: 651-53]. The ensuing period, accompanied by equally scanty settlement evidence, has become known as the ‘Archaic Gap'. The causes of this lacuna, which persists until the recrudescence of archaeological evidence in the Late Archaic period, remain unknown. Although various explanations have been forwarded, [@Coldstream1999: 301-302; @Huxley1994: 126; @Kotsonas2002: 41-44], there are now strong reasons to doubt the settlement was abandoned (Whitelaw, pers comm.).

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986l; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from Tomb 200+ and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.

For instance, Tomb 201 in the KNC contained a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123]. Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi; all contained imported items, were of ‘warrior grave’ type, and formed loci for later burials. He has argued that individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment, including the deposition of such items, in death [@Catling1995: 127–8].

Complementing the theme of heroes abroad has been a similar interest in foreigners at home. Two caches of gold jewellery, found in pits flanking the entrance to the Teke Tholos, were argued by Boardman [-@Boardman1967] to be foundation deposits – a Near Eastern custom – for the tomb of an immigrant Phoenician craftsman. Kotsonas has revisited this attribution, deeming the tholos more probably that of an elite individual or family with a monopoly over a metal workshop’s output [-@Kotsonas2006: 155–9]. While certainty in such individual cases may be moot, it seems unlikely that foreign traders or craftsmen never visited or resided at Knossos during the EIA [@Hoffman1997: 176–85; @Schreiber2003: 293-306].

@@

Generally speaking, the above analyses have tended to rely more on particular finds, tombs, or observable, but limited, patterns in the data; only a few authors have pursued more quantative approaches. Cluster analyses have been employed by both Cavanagh [-@Coldstream1996: 653-57] and Antoniadis [-@Antoniadis2012: 193-97], the most sophisticated statistical techniques so far applied to the material. Antoniadis [-@Antoniadis2012: 172-6] also divided the EIA tombs of Knossos into three groups, those with fewer than 10 pots, fewer than 50, and more than 50, and found that, in each successive case, the proportions of tombs containing 'oriental' imports and imitations increased, suggesting unequal access to such prestigious goods. Yet even here the available data could be further interrogated (with, for instance, the number of imported items *per burial*). The present paper thus seeks to extend and nuance a number of observations that have been made in previous studies, as well as revealing patterns thus far unacknowledged, through use of a more systematic approach to quantitative analysis.

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Variables includes the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016]. In total, this database comprises just over 4540 individual artefacts.

@@

Taken together, these databases make possible a diverse range of quantitative analyses. It must be conceded that, with many of the tombs evidently looted or damaged, this cannot be a complete picture of their original contents. If we assume, though, that natural or accidental destruction was relatively indiscriminate, and that, even in cases of looting, certain object classes (particularly cinerary urns) were seldom robbed, then the data are by no means rendered useless. As Snodgrass [-@Snodgrass1996] cautions, we should be extremely cautious about making negative arguments – that is, conclusions drawn about the *lack* of features in certain tombs – but may be somewhat more hopeful in the forming of positive ones.

@@

Ever more archaeologists have begun to advocate the utility, indeed the necessity, of making available not just the data, but the full methodologies underpinning our academic output. With the opportunities afforded by current technologies (digital databases, statistical software packages, online repositories etc.), there remain ever fewer obstacles to the full and free sharing of both data and workflows [see @article]. In Marwick's words [-@Marwick2017: 445], "[t]he technical problems are largely solved; the challenge now is to change the norms of the discipline to make high reproducibility a canonical attribute of high-quality scholarly work".

@@

There are, of course, valid reservations regarding issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research. But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit hugely from the greater accessability of datasets and workflows.

In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [see @Marwick2017; @inbook], an open-source package for the statistical software R Studio. This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the following analysis. Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R (and note also that, wherever possible, these are presented in colour-blind friendly palettes). This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and original article are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**). The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or perceived errors or inaccuracies in the data. It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.

@@

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen broadly follows that of Cavanagh [-@Cavanagh1996: 659-60], who took the number of pithoi (the most common urn) and the number of pithoi, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit (Figure \@ref(fig:burialestimate)). Here, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.

@@

Another major concern was the temporal uncertainty implicit in the assigning of individual vessels to ceramic periods. A Protogeometric B vessel could, on our best estimates, date anywhere between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data. A simple, if coarse, solution is to use the midpoint of the range assigned to each vessel (i.e. 825 for our PGB example). This is used in a couple of instances in the present paper, for assigning vessels to broad, century-long windows.

@@

But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi); the apparent cessation of depositions for 30 years in the 8th century is simply a fiction generated by the processing of the data.

A alternative approach, and the method employed extensively in this paper, is the use of so-called 'aoristic analysis' [for original use in criminology, see @Ratcliffe2000; for archaeological applications, see @Crema2010; @Crema2012; @Johnson2004a]. Here, rather than give a pot a single date, the entire time span under investigation is divided into equal chunks (in this case, decades), and probability values summing to one are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, datable between 840 and 810 BC, would register a value of 0.33 for each of those three decades. These values for individual vessels can then simply be summed to produce tomb- or cemetery-level signatures. A particular attraction of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" [@Johnson2004a, 450].

@@

As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations (see @Crema2010, 1123-24), they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.

@@

aorist_all_tomb_names$cemetery<-tombs_top_12$cem_condensed
